---
title: "Thursday Exam 1"
format: docx
editor: visual
---

```{r setup, include=FALSE}
library(dplyr)          ## Data Transformation 
library(ggplot2)        ## Data Visualization
library(GGally)         ## Data Visualization
set.seed(123)
```

### **Background**

You are working as a marketing analyst for a professional basketball team seeking to improve **ticket sales** through more effective promotional strategies. Over the course of a season, the team has run 100 marketing campaigns and recorded a variety of data points for each. These campaigns span digital outreach, physical advertising, and fan engagement tactics. Your job is to analyze the data and develop a regression model that helps predict total sales (`sales`) based on these different forms of marketing activity.

The variables collected for each campaign include: the number of `email_clicks` generated from newsletters, the number of `social_media_mentions` observed across platforms, the amount of `ad_spend` used on digital platforms, the number of `merch_giveaways` distributed during events, the estimated `billboard_exposure` in terms of views, and the number of `tv_spots` aired during peak hours. Using this dataset, you will build and compare several linear regression models to assess which combinations of marketing efforts most effectively predict ticket sales. This data was created based on simulated data.

### **Data Generation**

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Load necessary libraries

library(dplyr)          ## Data Transformation
library(tibble)         ## Data Transformation
library(tidyr)          ## Data Transformation
library(ggfortify)      ## Data Visualization
library(ggplot2)        ## Data Visualization
library(broom)          ## Data Analysis
library(lmtest)         ## Data Analysis 
library(car)            ## Data Analysis

source(here::here("assessment_regression.R"))

# ----------------------------
# 1) Set seed for reproducibility
# ----------------------------
set.seed(324)

# ----------------------------
# 2) Generate response: ticket_sales
# ----------------------------
n <- 1000
ticket_sales <- rnorm(n, mean = 500, sd = 55)  # avg revenue per campaign

# ----------------------------
# 3) Generate Model 1 and 2 predictors
# x1: email_clicks (moderately related to sales)
# x2: social_media_mentions (weakly related)
# x3: ad_spend (strongest relation + correlated with x2)
# ----------------------------
email_clicks <-  0.6 * ticket_sales + rnorm(n, mean = 0, sd = 20)
social_media_mentions <- 0.3 * ticket_sales + rnorm(n, mean = 0, sd = 70)
ad_spend <- 0.7 * ticket_sales + 0.5 * social_media_mentions + rnorm(n, mean = 0, sd = 100)

# ----------------------------
# 4) Generate Model 3 predictors (weak, independent)
# ----------------------------
merch_giveaways <- 0.6 * ticket_sales + rnorm(n, 0, 80)
billboard_exposure <- 0.5 * ticket_sales + rnorm(n, 0, 30)
tv_spots <- 0.2 * ticket_sales + rnorm(n, 0, 25)

# ----------------------------
# 5) Create DataFrame
# ----------------------------
marketing_df <- data.frame(
  ticket_sales = round(ticket_sales, 2),
  email_clicks = round(email_clicks, 2),
  social_media_mentions = round(social_media_mentions, 2),
  ad_spend = round(ad_spend, 2),
  merch_giveaways = round(merch_giveaways, 2),
  billboard_exposure = round(billboard_exposure, 2),
  tv_spots = round(tv_spots, 2)
)


```

```{r}


# Calculate min and max for each variable
marketing_df %>%
  summarise(
    across(everything(), list(min = min, max = max), .names = "{.col}__{.fn}")
  ) %>%
  pivot_longer(cols = everything(),
               names_to = c("variable", "stat"),
               names_sep = "__") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(variable)
```

### **Model Building**

```{r}
## Model 1
model1 <- lm(ticket_sales ~ email_clicks + social_media_mentions, data = marketing_df)

## Model 2
model2 <- lm(ticket_sales ~ email_clicks + social_media_mentions + ad_spend, data = marketing_df)

## Model 3
model3 <- lm(ticket_sales ~ merch_giveaways + billboard_exposure + tv_spots, data = marketing_df)
```

```{r}
model_1_summary <- summarize_reg_model(model1,"Model 1")
model_2_summary <- summarize_reg_model(model2,"Model 2")
model_3_summary <- summarize_reg_model(model3,"Model 3")

bind_rows(
  model_1_summary,
  model_2_summary,
  model_3_summary
) 
```

Note:

Model 1 - `ticket_sales ~ email_clicks + social_media_mentions`

Model 2 - `ticket_sales ~ email_clicks + social_media_mentions + ad_spend`

Model 3 - `ticket_sales ~ merch_giveaways + billboard_exposure + tv_spots`

\newpage

**An employee at the company suggested that model 2 should be used for predicting ticket sales.**

```{r}
summary(model2)
```

```{r}
anova(model1,model2)
```

```{r}
## Perform residual analysis plot
autoplot(model2) + ## from ggfortify
  theme_bw()
```

```{r}
resettest(model2)                # From lmtest package
dwtest(model2)                   # From lmtest package
shapiro.test(resid(model2))      # Base R
bptest(model2)                   # From lmtest package
```

**Model 2: VIF**

```{r}
vif(model2)   ## From car package
```

**Model 2: Partial Plots**

```{r}
avPlots(model2)       ## From car package
```
