---
title: "Friday Quiz 2"
format: pdf
editor: visual
---

# School Data

This synthetic dataset simulates data collected from a high school’s academic enrichment program, designed to explore how various student habits relate to academic success. In particular, we imagine researchers created three survey-based variables: **x1**, representing weekly hours spent in *peer-led study groups*; **x2**, weekly hours dedicated to *quiet individual study time*; and **x3**, weekly hours spent in *school-sponsored extracurricular clubs*. The outcome variable, **y**, represents the student’s final *academic performance index* for the term, scaled from 0 to 100.

The data was generated in such a way that the three predictors are close to statistically independent (i.e., there's no strong correlation between them), allowing for a clean interpretation of each variable’s unique contribution to academic performance. For example, a student could study independently without necessarily being involved in peer groups or clubs, making these behaviors distinct. In this synthetic model, we assume that **individual study (x2)** has the strongest positive association with performance, **peer groups (x1)** have a moderate positive effect, and **extracurricular clubs (x3)** show a slight negative association—perhaps reflecting a trade-off in time. The goal of this simulation is to help students and teachers analyze how various independent efforts and activities can affect outcomes when not confounded by overlapping behaviors.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)                ## Data Transformation
library(ggplot2)              ## Data Visualization
library(GGally)               ## Data Visualization
library(MASS)                 ## Multivariate Normal Distribution

set.seed(123)

sample_size <- 100

# Define realistic school-based means
real_means <- c(3, 7, 5)  # study_group, self_study, clubs

# Independent variables with identity covariance
X <- mvrnorm(n = sample_size, mu = real_means, Sigma = diag(3))

# Response: y = 3 + 1*x1 + 2*x2 - 1*x3 + noise
y <- 3 + X %*% c(1, 2, -1) + rnorm(sample_size) + 60


# Build and round the data frame
school_df  <- as.data.frame(bind_cols(y, X))
colnames(school_df ) <- c("performance", "study_group", "self_study", "clubs")
school_df <- round(school_df , 2)
```

```{r echo=FALSE}
summarize_reg_model <- function(model,model_description) {
  # Get number of observations and predictors
  n <- length(residuals(model))
  p <- length(coefficients(model)) - 1
  
  # 0) RSS 
  RSS <- round(sum(residuals(model)^2),2)
  
  # 1) RSE
  RSE <- round(sqrt(RSS / (n - p - 1)),2)
  
  # 2) R-Squared and Adjusted R-Squared
  R2 <- round(summary(model)$r.squared,2)
  adj_R2 <- round(summary(model)$adj.r.squared,2)
  
  # 3) AIC
  aic_val <- round(AIC(model),2)
  
  # 4) BIC
  bic_val <- round(BIC(model),2)
  
  # Create summary dataframe
  mlr_metrics <- data.frame(
    type = model_description,
    RSS = RSS,
    RSE = RSE,
    R2 = R2,
    Adj_R2 = adj_R2,
    AIC = aic_val,
    BIC = bic_val
  )
  
  return(mlr_metrics)
}
```

## Review of Model Metrics

```{r echo=FALSE}
# Fit multiple regression models
school_model_1 <- lm(performance ~ study_group, data = school_df)
school_model_2 <- lm(performance ~ study_group + self_study, data = school_df)
school_model_3 <- lm(performance ~ study_group + self_study + clubs, data = school_df)

# If you have a summarize_reg_model() function
summary_model_1 <- summarize_reg_model(school_model_1,
                                       "Model 1: Performance ~ Study Group")
summary_model_2 <- summarize_reg_model(school_model_2,
                                       "Model 2: Performance ~ Study Group + Self Study")
summary_model_3 <- summarize_reg_model(school_model_3,
                                       "Model 3: Performance ~ Study Group + Self Study + Clubs")

# Combine summaries
summary_df <- bind_rows(summary_model_1, summary_model_2, summary_model_3)

# Display results
summary_df %>%
  knitr::kable() %>% 
  ## Hold Position
  kableExtra::kable_styling(latex_options = c("hold_position", "scale_down"),full_width = FALSE, font_size = 10) 
  
  

```

\newpage

# Model Comparisons

## Model Comparison 1

```{r}
anova(school_model_1, school_model_2)
```

## Model Comparison 2

```{r}
anova(school_model_2, school_model_3)
```

\newpage

# Review of Selected of Final Model

```{r}
summary(school_model_3)
```
