---
title: "Homework 1"
format: pdf
editor: visual
---

The `mtcars` dataset is a classic and widely used dataset in R that contains specifications and performance data for 32 different car models from the 1974 Motor Trend magazine. Each row represents a unique vehicle, and each column records a specific attribute related to engine performance, design, or efficiency. Some key variables include `mpg` (miles per gallon), `hp` (gross horsepower), `wt` (weight in 1000 lbs), `drat` (rear axle ratio), and `qsec` (quarter-mile time). Additionally, the dataset includes categorical variables encoded as numeric values, such as `cyl` (number of cylinders), `am` (transmission type), and `gear` (number of forward gears).

This dataset is frequently used in regression modeling and statistical learning due to its compact size, real-world relevance, and mixture of quantitative and categorical variables. Analysts often model fuel efficiency (`mpg`) as a function of other variables to understand how engine power, vehicle weight, or gear ratios impact gas mileage. With its balance of complexity and interpretability, `mtcars` serves as a great playground for developing skills in exploratory data analysis, model selection, variable interpretation, and diagnostics in both teaching and applied settings.

```{r message=FALSE, warning=FALSE, echo=FALSE}

library(dplyr)            ## Data Transformation   --- T
library(ggplot2)          ## Data Visualization    --- V
library(GGally)           ## Data Visualization    --- V

```

\newpage

## ðŸ” Exploratory Visualization

```{r message=FALSE, warning=FALSE}
mtcars %>% 
  select(mpg, hp, wt, drat) %>% 
  ggpairs()
```

\newpage

## ðŸ“ˆ Model 1: `mpg ~ hp`

```{r}
model1 <- lm(mpg ~ hp, data = mtcars)
summary(model1)
```

```{r echo=FALSE}
 
# Metrics
n <- nrow(mtcars)
p <- length(coef(model1)) - 1
RSS <- sum(residuals(model1)^2)
RSE <- sqrt(RSS / (n - p - 1))
R2 <- summary(model1)$r.squared
adj_R2 <- summary(model1)$adj.r.squared
aic_val <- AIC(model1)
bic_val <- BIC(model1)

model1_metrics <- data.frame(
  type = "Model 1: mpg ~ hp",
    RSS = round(RSS, 2),
  RSE = round(RSE, 2),
  R2 = round(R2, 2),
  Adj_R2 = round(adj_R2, 2),
  AIC = round(aic_val, 2),
  BIC = round(bic_val, 2)
) 
model1_metrics
```

------------------------------------------------------------------------

\newpage

## ðŸ“ˆ Model 2: `mpg ~ hp + wt`

```{r}
model2 <- lm(mpg ~ hp + wt, data = mtcars)
summary(model2)
```

```{r echo=FALSE}
# Metrics
p <- length(coef(model2)) - 1
RSS <- sum(residuals(model2)^2)
RSE <- sqrt(RSS / (n - p - 1))
R2 <- summary(model2)$r.squared
adj_R2 <- summary(model2)$adj.r.squared
aic_val <- AIC(model2)
bic_val <- BIC(model2)

model2_metrics <- data.frame(
  type = "Model 2: mpg ~ hp + wt",
    RSS = round(RSS, 2),
  RSE = round(RSE, 2),
  R2 = round(R2, 2),
  Adj_R2 = round(adj_R2, 2),
  AIC = round(aic_val, 2),
  BIC = round(bic_val, 2)
)
model2_metrics
```

```{r}
# Partial F-test: Does adding wt help?
anova(model1, model2)
```

------------------------------------------------------------------------

\newpage

## ðŸ“ˆ Model 3: `mpg ~ hp + wt + drat`

```{r}
model3 <- lm(mpg ~ hp + wt + drat, data = mtcars)
summary(model3)
```

```{r echo=FALSE}
# Metrics
p <- length(coef(model3)) - 1
RSS <- sum(residuals(model3)^2)
RSE <- sqrt(RSS / (n - p - 1))

R2 <- summary(model3)$r.squared
adj_R2 <- summary(model3)$adj.r.squared
aic_val <- AIC(model3)
bic_val <- BIC(model3)

model3_metrics <- data.frame(
  type = "Model 3: mpg ~ hp + wt + drat",
    RSS = round(RSS, 2),
  RSE = round(RSE, 2),
  R2 = round(R2, 2),
  Adj_R2 = round(adj_R2, 2),
  AIC = round(aic_val, 2),
  BIC = round(bic_val, 2)
)
model3_metrics
```

```{r}
# Partial F-test: Does adding drat help?
anova(model2, model3)
```

------------------------------------------------------------------------

\newpage

## Model Comparison Summary

```{r}
bind_rows(model1_metrics, model2_metrics, model3_metrics)
```
