---
title: "Final Exam STAT 324"
format: docx
editor: visual
---

## 1. Introduction & Research Goal

In the dataset provided, the response variable is `sfood` which represents the family income in British pounds per week. `sfood`—the household share of total expenditure devoted to food—is a classic Engel-curve outcome: it naturally varies with purchasing power (income) and household composition (kids, age group), making it an intuitive, policy-relevant response for modeling how spending priorities shift across socioeconomic strata. This variable serves as the response variable in the analysis, indicating it’s the outcome that the analysis seeks to explain or predict. The dataset includes a range of explanatory variables, each representing a different share of weekly expenditures as a proportion of the total. These include sfuel (share of fuel expenditures), sclothes (share of clothing expenditures), salcohol (share of alcohol expenditures), stransport (share of transportation expenditures), and sother (share of other expenditures). These variables provide insight into how household spending is distributed across various categories.

Additionally, the dataset contains variables that capture the total expenditure (totexpend) of a household in British pounds per week, the age of the household head (age), and the number of children in the household (kids), which is categorized as either 1 or 2. These variables can be used to understand the demographic and economic factors that might influence the family income. The inclusion of ltotexpend and lincome, which are the natural logarithms of total expenditure and income respectively, suggests that the relationship between the logarithm of income and expenditures could be explored. Similarly, agesq, the square of the age of the household head, might be used to investigate non-linear relationships between age and income.

The variable `high_sfuel` is a binary variable that categorizes the share of fuel expenditures (`sfuel`) into "high" if it's greater than the median, and "low" otherwise. This variable provides a simple way to compare households with high and low fuel expenditures. The variable `kids_binary` is another binary variable that categorizes the number of children (`kids`) into "one" if there is one child, and "two" otherwise. This variable simplifies the analysis by reducing the number of children to two categories.

The variable `sclothes_group` categorizes the share of clothing expenditures (`sclothes`) into three groups: "low", "medium", and "high". This variable allows for an analysis of the impact of different levels of clothing expenditures on the response variable. The variable `age_group` categorizes the age of the household head (`age`) into three groups: "young", "middle-aged", and "old". This variable allows for an analysis of the impact of different age groups on the response variable. `stransport` measures the fraction of a household’s total outlays devoted to transportation. A larger `stransport` value means transportation absorbs a bigger slice of the family budget. Finally, the variable `salcohol_group` categorizes the share of alcohol expenditures (`salcohol`) into four groups: "low", "medium", "high", and "very high". This variable allows for an analysis of the impact of different levels of alcohol expenditures on the response variable. Overall, this dataset provides a comprehensive view of household finances, with a focus on how various factors relate to family income.

\newpage

```{r message = FALSE, warning = FALSE}
library(dplyr)                ## Data Transformation    --- T
library(ggplot2)              ## Data Visualization     --- V
library(ggfortify)            ## Data Visualization     --- V
library(MASS)                 ## Data Analysis          --- A
library(leaps)                ## Data Analysis          --- A
library(lmtest)               ## Data Analysis          --- A
library(broom)                ## Data Analysis          --- A
library(car)                  ## Data Analysis          --- A
```

## 2. Data Description

Briefly describe the 1,519-household dataset, noting each key variable (shares, totals, logs, groups) and highlighting categorical groupings created for fuel, clothing, alcohol, kids, and age. Emphasize that these groupings enable flexible modeling of nonlinear or subgroup-specific patterns.

```{r message = FALSE, warning = FALSE}
#| echo: false
load("../Data/expendshares.RData")

expend_df <- data
```

## 3. Data Preparation

Outline the cleaning and transformation steps: creation of binary and multi-level factors, grand-mean and group-mean centering, and polynomial terms. Note that these steps reduce multicollinearity and prepare predictors for interaction analyses.

```{r message = FALSE, warning = FALSE}
mod_1_expend_df <- expend_df %>%
  mutate(high_sfuel = if_else(sfuel > median(sfuel), "high", "low")) %>% 
  ## mutate binary variable for kids
  mutate(kids_binary = if_else(kids == 1, "one", "two")) %>%
  ## mutate sclothes into three groups
  mutate(sclothes_group = case_when(
    sclothes < 0.08 ~ "low",
    sclothes >= 0.08 & sclothes < 0.2 ~ "medium",
    sclothes >= 0.2 ~ "high"
  )) %>%
  ## mutate age into three groups
  mutate(age_group = case_when(
    age < 30 ~ "young",
    age >= 30 & age < 50 ~ "middle-aged",
    age >= 50 ~ "old"
  )) %>% 
  ## mutate share alcoholo intp four groups
  mutate(salcohol_group = case_when(
    salcohol < 0.1 ~ "low",
    salcohol >= 0.1 & salcohol < 0.2 ~ "medium",
    salcohol >= 0.2 & salcohol < 0.3 ~ "high",
    salcohol >= 0.3 ~ "very high"
  ))
```

```{r message = FALSE, warning = FALSE}

mod_2_expend_df <- mod_1_expend_df %>% 
  group_by(sclothes_group) %>% 
  mutate(totexpend_gc = totexpend - mean(totexpend, na.rm = TRUE)) %>% 
  ungroup() %>% 
  mutate(stransport_c = stransport - mean(stransport, na.rm = TRUE))  %>% 
  mutate(totexpend_c = totexpend - mean(totexpend, na.rm = TRUE))  
```

\newpage

## 4. Exploratory Visualizations

List the three plots: (i) quadratic sfood–totexpend curve, (ii) stratified lines by sclothes_group, and (iii) group-mean-centered lines. Explain that these visuals reveal curvature, group-specific slopes, and within-group effects that motivate the formal models.

```{r}
ggplot(mod_1_expend_df, aes(x = totexpend, y = sfood)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, linewidth = 1.2) +
  labs(
    title = "Food-Expenditure Share vs. Total Expenditures",
    x = "Total Expenditures (totexpend)", y = "share of food (sfood)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

\newpage

```{r}
ggplot(mod_1_expend_df, aes(x = totexpend, y = sfood, colour = sclothes_group)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
    labs(
    title  = "Food-Share vs. Total Expenditure by\n Clothing-Spend Group",
    x      = "Total Expenditure (totexpend)",
    y      = "Share of Food Expenditure (sfood)",
    colour = "Clothing-Spend Group"
  )  +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

\newpage

```{r}
ggplot(mod_2_expend_df, aes(totexpend_gc, sfood, colour = sclothes_group)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.1) +
 labs(
    title  = "Food-Share vs. Group-Mean-Centered\n Total Expenditure",
    x      = "Total Expenditure (group-mean centered, totexpend_gc)",
    y      = "Share of Food Expenditure (sfood)",
    colour = "Clothing-Spend Group"
  )+
  theme_minimal()
```

\newpage

## 5. Investigation into Hypothesized Models

| Model \# | Formula (new data) | What it captures |
|----|----|----|
| **1** | `sfood ~ totexpend + stransport` | Two quantitative predictors (`totexpend`, `stransport`). |
| **2** | `sfood ~ totexpend + sclothes_group` | Quantitative predictor (`totexpend`) plus a categorical grouping (`sclothes_group`). |
| **3** | `sfood ~ totexpend * stransport` | Interaction between two quantitative variables (does the food-share response to total spending vary with transportation share?). |
| **4** | `sfood ~ totexpend * sclothes_group` | Interaction between a quantitative predictor and a categorical group (does the slope for total spending differ across clothing-spend groups?). |
| **5** | `sfood ~ totexpend_c * stransport_c` | Mean-centered versions of both quantitative predictors plus their interaction (easier interpretation, lower multicollinearity). |
| **6** | `sfood ~ totexpend + I(totexpend^2)` | Polynomial term in `totexpend` to allow curvature in the Engel-type relationship. |

\newpage

```{r}
options(scipen = 9)
# Model 1: two quantitative predictors
model_1 <- lm(sfood ~ totexpend + stransport, data = mod_2_expend_df)
summary(model_1)
```

\newpage

```{r}
# Model 2: quantitative + categorical
model_2 <- lm(sfood ~ totexpend + sclothes_group, data = mod_2_expend_df)
summary(model_2)
```

\newpage

```{r}
# Model 3: interaction between two quantitative variables
model_3 <- lm(sfood ~ totexpend * stransport, data = mod_2_expend_df)
summary(model_3)
```

\newpage

```{r}
# Model 4: interaction between quantitative and categorical predictor
model_4 <- lm(sfood ~ totexpend * sclothes_group, data = mod_2_expend_df)
summary(model_4)
```

\newpage

```{r}
# Model 5: centered variables with interaction
model_5 <- lm(sfood ~ totexpend_c*stransport_c, data = mod_2_expend_df)
summary(model_5)
```

\newpage

```{r}
# Model 6: polynomial (quadratic) term for lincome
model_6 <- lm(sfood ~ totexpend + I(totexpend^2), data = mod_2_expend_df)
summary(model_6)
```

\newpage

## 6. Model Selection Procedures

### a. Stepwise Regression

Explain the stepwise AIC search and best-subsets (adjusted-R\^2) approach run on an expanded predictor set. Note that these automated tools cross-check the hand-built models and suggest a parsimonious final candidate.

```{r}
# Start with a full model
full_model <- lm(sfood ~ totexpend  + stransport + sclothes + income + age, data = mod_2_expend_df)

# Stepwise model using both directions (backward & forward)
step_model <- stepAIC(full_model, direction = "both", trace = TRUE)

# Summary of final selected model
summary(step_model)
```

#### i. Multicollinearity Check

```{r}
car::vif(step_model)
```

\newpage

#### ii. Outliers, Leverage, and Influence

Note that standardized residuals, Cook’s distance, and leverage values are extracted and the most influential 15 cases tabulated; any problematic observations are flagged for further scrutiny.

```{r}
# Add standardized residuals
augmented_df <- augment(step_model) ## from broom package

rownames(augmented_df) <- NULL

augmented_df %>% 
  dplyr::select( -totexpend, -stransport, -sclothes, -income, -age,-sfood) %>%
  arrange(desc(.cooksd)) %>% 
  head(15) %>% 
  round(3)
```

\newpage

#### iii. Diagnostics & Assumption Tests

List the residual tests performed (RESET for functional form, Durbin-Watson for autocorrelation, Shapiro for normality, Breusch-Pagan for heteroskedasticity). State that plots and p-values guide whether assumptions hold or require remediation.

```{r}
autoplot(step_model) + 
  theme_bw()
```

\newpage

```{r}
resettest(step_model)                # From lmtest package
dwtest(step_model)                   # From lmtest package
shapiro.test(resid(step_model))      # Base R
bptest(step_model)                   # From lmtest package
```

\newpage

### b. Model Selection Best Subset

```{r}
# Fit all subsets model
subset_model <- regsubsets(
  sfood ~ totexpend  + stransport + sclothes + income + age, data = mod_2_expend_df,
  nvmax = 5
)

plot(x = subset_model,        ## Specify Model Fit
     scale = "adjr2"          ## Specify Selection Metric
)
```
